Ages of Planets and Stars Observed by Kepler

### 2024-04-16 paper/ms.tex
update ms

%%%%%%%%%%%%%
% OFT-CITED %
%%%%%%%%%%%%%
\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}

@@

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.
In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Many of
these reported detections have lower variability amplitudes, and
longer periods, than those reported by
\citetalias{McQuillan_2014}.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.  Reviewing the resulting period detections, we
learned that 32 KOIs with orbital and rotation periods within
$\approx$20\% were excluded from the \citetalias{Santos_2019} and
\citetalias{Santos_2021} catalogs (A.~Santos, priv.~comm.) ; we choose
to include these objects.\footnote{In general, excluding stars in this
category would impose a bias against young close-in planets, due to
the natural commensurability between the rotation rates of young
stars, and Kepler's increased sensitivity to short-period objects.
While some of these KOIs are known false positives, our analysis
naturally excludes such objects at later steps.  As one motivating
example, this subset of stars happens to include {\it
e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
based on membership in RSG-5 \citep{Bouma_2022b}.  }
The final tally of stars from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our corrected KOI list is
\nuniqstarsantosrot\ Kepler stars with rotation periods.

@@

		\subfloat{
		\includegraphics[width=0.5\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf todo}.
	}
	\label{fig:camd}
\end{figure*}

@@

As our default source for stellar temperatures and surface gravities,
we adopt the values from the Gaia-Kepler Stellar Properties Catalog
(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
available for stars with ``AAA'' 2MASS photometry, measured parallaxes
in Gaia DR2,  and $g$-band photometry from either the KIC or the
Kepler-INT survey.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids
\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
map.  The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
stars with rotation periods.

@@

For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities, we adopt the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog, and are mostly derived from photometry.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
from DR25.

As a sanity check on our effective temperature scale, we then used the 
the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
relation suggested by \citet{Curtis_2020}.
This latter effective temperature scale was demonstrated to have a 
scatter of $\approx$50\,K for stars near the main-sequence.
We find {\bf X, Y, Z.}
We ultimately favor the \citet{Berger_2020a_catalog} scale
because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
isochronal fit.

@@

\subsection{Star Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select the subset stars for which we expect these age to in fact be
valid, we then build a set of quality flags.  When and
how these flags should be applied depends on the question being asked.
If the goal is to construct a false-positive free sample, they might
all be applied.
However if the goal is constructing a complete sample of young stars,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  
The flags in Tables~\ref{tab:stars} and~\ref{tab:planets} enable
selecting for such cases using bitflags.

{\it Subgiants}---We flagged stars as possible subgiants if they had
$\log g < 4.2$.  We set this threshold by examining the
relevant flag from \citet{berger_2018_radii_evolnstates},
cross-matched against \citet{Berger_2020a_catalog}.
\citet{berger_2018_radii_evolnstates} performed this classification
using the PARSEC evolutionary models, stellar radii, and DR25 stellar
properties catalog effective temperatures \citep{Mathur_2017}.

{\it Photometric binaries}---We plotted color--absolute magnitude
diagrams using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
loci to flag over or under-luminous stars in each diagram, typically
more than $\approx$0.5 magnitudes above or below the main sequence.
%TODO refer to a figure!!!

{\it RUWE}---Following recommendations in the Gaia DR3
documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
we flagged stars with a Gaia DR3 renormalized unit weight error (RUWE)
exceeding 1.4 as possible binaries.  Such astrometric outliers can be
either bonafide astrometric binaries, or more often are marginally
resolved point sources for which the single-source PSF model used in
the Gaia pipeline induces spurious astrometric scatter.

{\it Crowding}---We searched the Gaia DR3 point source catalog for
stars within 1 Kepler pixel (4$''$) of every target star.
While many such companions are not physically associated with the
target star, their presence can bias stellar parameters and stellar
rotation period measurements.
We therefore flag any stars with neighbors down to 10$\times$ the 
brightness of the target star within this region ($\Delta G < 2.5$).

{\it Gaia DR3 Non-Single-Stars}---The Gaia DR3
\texttt{non\_single\_star} column in the \texttt{gaia\_source} table
flags known eclipsing, astrometric, and spectroscopic binaries.  We
directly include this column.

@@

Kepler~411 is also odd, due to the 58\,day-outer planet.
\citet{2019A&A...624A..15S} conducted a transiting timing analysis of the
system, confirming Kepler-411d and argued for the presence of a fourth
non-transiting planet, Kepler-411e, based on the TTVs seen in Kepler-411d and
assuming that the system would be nearly coplanar.
Regarding the stellar age, \citet{2019A&A...624A..15S} used the 
\citet{2007ApJ...669.1167B} relation, which yielded an estimate of 212$\pm$31\,Myr.
However, the star's temperature ($\approx$4900\,K) puts it in a regime where
the Barnes~relation has been known to not fit observations for quite some time
\citep[see][Fig.~9]{2008ApJ...687.1264M}.
Our adopted age of $\approx$780$\pm$170\,Myr agrees with the open cluster
calibrators.

>>>>

%%%%%%%%%%%%%
% OFT-CITED %
%%%%%%%%%%%%%
\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}
\defcitealias{Fulton_2018}{F18}
\defcitealias{Berger_2020a_catalog}{B20}

@@

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.
In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Many of
these reported detections have lower variability amplitudes, and
longer periods, than those reported by
\citetalias{McQuillan_2014}.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars} {\bf TODO make this}.  Reviewing the resulting period detections, we
learned that 32 KOIs with orbital and rotation periods within
$\approx$20\% were excluded from the \citetalias{Santos_2019} and
\citetalias{Santos_2021} catalogs (A.~Santos, priv.~comm.) ; we choose
to include these objects.\footnote{In general, excluding stars in this
category would impose a bias against young close-in planets, due to
the natural commensurability between the rotation rates of young
stars, and Kepler's increased sensitivity to short-period objects.
While some of these KOIs are known false positives, our analysis
naturally excludes such objects at later steps.  As one motivating
example, this subset of stars happens to include {\it
e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
based on membership in RSG-5 \citep{Bouma_2022b}.  }
The final tally of stars from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our corrected KOI list is
\nuniqstarsantosrot\ Kepler stars with rotation periods.

@@

		\subfloat{
		\includegraphics[width=0.5\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf todo}.
	}
	\label{fig:stellarprops}
\end{figure*}

@@

As our default source for stellar temperatures and surface gravities,
we adopt the values from the Gaia-Kepler Stellar Properties Catalog
(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
available for stars with ``AAA'' 2MASS photometry, measured parallaxes
in Gaia DR2,  and $g$-band photometry from either the KIC or the
Kepler-INT survey.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids
\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
map.  The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
stars with rotation periods.
\citet{David_2021} compared the stellar parameters ($T_{\rm eff}$,
$R_\star$, [Fe/H]) derived from the \citet{Berger_2020a_catalog}
isochronal approach with the spectroscopic parameters from
\citet{Fulton_2018}.  While this approach is one way of validating the
\citet{Berger_2020a_catalog} output, the temperature scales did show a
few-percent difference, with \citetalias{Fulton_2018} quoting higher
temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
and lower temperatures for early F dwarfs.

@@

For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities from
\citetalias{Berger_2020a_catalog}, we adopt the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog, and are mostly derived from photometry.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
from DR25.

%TODO: the relevant check here is Berger20 teff vs CKS spec teff...
%  As a sanity check on our effective temperature scale, we then used the 
%  the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
%  BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
%  relation suggested by \citet{Curtis_2020}.
%  This latter effective temperature scale was demonstrated to have a 
%  scatter of $\approx$50\,K for stars near the main-sequence.
%  We find {\bf X, Y, Z.}
%  We ultimately favor the \citet{Berger_2020a_catalog} scale
%  because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
%  isochronal fit.

@@

\subsection{Gyrochronology Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select the subset stars for which we expect these age to in fact be
valid, we then build a set of quality flags.  When and
how these flags should be applied depends on the question being asked.
If the goal is to construct a false-positive free sample, they might
all be applied.
If the goal is constructing a complete sample of young stars,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  
The flags in Tables~\ref{tab:stars} and~\ref{tab:planets} enable
selecting for such cases using bitflags.

{\it Temperature range (bit 0)}---We required stars to have effective
temperatures $T_{\rm eff}$ between 3800 and 6200\,K
\citep{Bouma_2023}.   Stars hotter than 6200\,K spin down very slowly,
if at all, limiting the utility of rotation-based ages.  Stars cooler
than 3800\,K do seem to spin down over gigayear timescales (e.g. CITE
E.Newton thesis, the Engle paper, and Chiti's thing), but the
currently available open cluster data have yet to clarify at what
times, if any, the intrinsic scatter in the population decreases.

{\it Surface gravity (bit 1)}---We flagged stars as possible subgiants if they had
$\log g < 4.2$.  We set this threshold by examining the
relevant flag from \citet{berger_2018_radii_evolnstates},
cross-matched against \citet{Berger_2020a_catalog}.
\citet{berger_2018_radii_evolnstates} performed this classification
using the PARSEC evolutionary models, stellar radii, and DR25 stellar
properties catalog effective temperatures \citep{Mathur_2017}.

{\it Absolute luminosity (bit 2)}---We calculated the absolute Gaia DR3
$G$-band luminosity, ignoring reddening, using the reported apparent
$G$-band magnitude and parallaxes.
We flag stars with $M_{\rm G} < 3.9$, corresponding to main-sequence
spectral types earlier than $\approx$F8V \citep{Pecaut_2013}.

{\it Known eclipsing binaries (bit 3)}---We flagged any stars reported to be
in the final Kepler eclipsing binary catalog (KEBC;
\citealt{2016AJ....151...68K}).

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In instances
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun/}; we
used 4$''$ rather than the 1$''$ match because 91 KOIs are omitted in
the 1$''$ match, 23 of which have either ``candidate'' or
``confirmed'' status.  Almost all of these 23 are M dwarfs with high
proper motions.  Performing the same crossmatch at 4 arcseconds, only
two KOIs (KOI-3993 and KOI-6531), both known false positives, fail to
yield Gaia DR3 crossmatches.}, which is a crossmatch of the NASA
Exoplanet Archive \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia
DR3.  For KIC stars with multiple potential Gaia matches within the
4$''$ search radius, we adopted the brightest star as the default
match.  in most such cases this was unambiguous because there is a
large brightness difference between the primary and any apparent
neighbors.  Cases with multiple stars within 4$''$ within $\Delta
G$$<$$0.5$\,mag are noted using bit flag 5.  The separation
distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
of candidate matches are within 1$''$.   We nonetheless noted an
upturn in the candidate match rate from 3-4$''$; such sources are
flagged using bit 4.

{\it Gaia DR3 non-single-stars (bit 6)}---The Gaia DR3
\texttt{non\_single\_star} column in the \texttt{gaia\_source} table
flags known eclipsing, astrometric, and spectroscopic binaries.  We
directly include this column.

{\it RUWE (bit 7)}---Following recommendations in the Gaia DR3
documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
we flagged stars with a Gaia DR3 renormalized unit weight error (RUWE)
exceeding 1.4 as possible binaries.  Such astrometric outliers can be
either bonafide astrometric binaries, or more often are marginally
resolved point sources for which the single-source PSF model used in
the Gaia pipeline induces spurious astrometric scatter.

{\it Crowding (bit 8)}---We searched the Gaia DR3 point source catalog
for stars within 1 Kepler pixel (4$''$) of every target star.  While
many such companions are not physically associated with the target
star, their presence can bias stellar parameters and stellar rotation
period measurements.  We therefore flag any stars with neighbors down
to 10$\times$ the brightness of the target star within this region
($\Delta G < 2.5$).

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in color--absolute
magnitude diagrams, and in diagrams of e.g.~surface gravity as a
function of temperature (Figure~\ref{fig:stellarprops}).  This poses a
problem for rotation-based age-dating because {\it i)} ``stars'' far
from the main-sequence include unresolved and unidentified photometric
binaries, and {\it ii}) rotation periods of single stars far from the
main-sequence have likely evolved due to effects other than magnetic
braking.  The added complications of reddening and metallicity make
imposing a simple cut in these diagrams challenging.  After exploring
various options, we found that the $+1$$\sigma$ relative isochronal
uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{\rm
B20,iso}/t_{\rm B20,iso}$, traced the relevant region of parameter
space quite well.
Another way of saying this is that gyrochronology is likely to be less
reliable for stars for which the upper limit on the relative
isochronal uncertainty is small.

%FIXME TODO LEFT OFF HERE...

df['b20t2_rel_E_Age'] = np.abs(df['b20t2_E_Age'])/df['b20t2_Age']

This is potentially an issue for rotation-based age dating,
particularly

using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
loci to flag over or under-luminous stars in each diagram, typically
more than $\approx$0.5 magnitudes above or below the main sequence.
%TODO refer to a figure!!!

@@

Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
transiting planets at 3.0, 7.8, and 58\,days, has a peculiar
architecture.  \citet{2019A&A...624A..15S} conducted a transiting
timing analysis of the system, confirming Kepler-411d and arguing for
the presence of a fourth non-transiting planet, Kepler-411e, based on
the TTVs seen in Kepler-411d and assuming that the system is nearly
coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
the \citet{2007ApJ...669.1167B} relation and the measured rotation
period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
star's temperature ($\approx$4900\,K) puts it in a regime where the
\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
observations for quite some time
\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.

Kepler-63b is interesting in that it is {\it missing} from our
catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
a polar orbit based on both the Rossiter-McLaughlin effect and a
starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
qualitatively consistent with the estimates from
\citet{2013ApJ...775...54S}, though with larger uncertainties because
our age accounts for the intrinsic scatter in the open cluster
calibration sequences \citep{Bouma_2023}.  
While the usual catalogs (\citetalias{Santos_2019},
\citetalias{Santos_2021}, \citetlias{McQuillan_2014}, and
\citetalias{Mazeh_2015}) do not report this star's rotation period.
\citetalias{Santos_2021} flagged the system as ``potentially'' having
rotation-based modulation.
{\bf todo: how many comparable missed cases are there?}.
%TODO FIXME it's bc you need to do mcquillan 2013....
%
%... David+2021 used:
% Angus+2018... McQuillan+2013... Mazeh+2015... Walkowicz&Basri2013...
